---
title: "Task Code — Microbiology: Bacterial Growth Curve Analysis"
author: "Ange Maxime TCHOUTANG"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    number_sections: true
---

```{r setup, include=FALSE}
# Global chunk options:
# - echo    : show code in output
# - message : suppress package loading messages
# - warning : suppress warnings from plots/stats
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 12,
  fig.height = 7
)
```

---

# 1. Load Packages

```{r packages}
# pacman checks if packages are installed; installs missing ones then loads all
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,  # data wrangling (dplyr, tidyr) + ggplot2 for plotting
  ggplot2,    # explicit load for clarity
  patchwork,  # combine multiple ggplots into one panel
  scales,     # axis formatting helpers
  ggpubr      # adds stat_compare_means() for easy p-value display on plots
)
```

---

# 2. Load & Inspect the Data

```{r load-data}
# ── Load the TSV dataset directly from GitHub ────────────────
url <- "https://raw.githubusercontent.com/HackBio-Internship/2025_project_collection/refs/heads/main/Python/Dataset/mcgc.tsv"

mcgc_raw <- read.delim(url, sep = "\t", stringsAsFactors = FALSE)

# ── Quick overview ───────────────────────────────────────────
dim(mcgc_raw)       # rows = time points, cols = 1 time + 36 wells
head(mcgc_raw[, 1:10])  # preview first 10 columns
```

> **Observation:** The dataset has **time** as the first column (in minutes),
> followed by **36 well columns** named A1–A12, B1–B12, C1–C12.
> Each well contains OD600 readings — a measure of bacterial cell density.
> Rows = time points measured every 15 minutes.

---

# 3. Understanding the Plate Layout (Metadata)

```{r metadata}
# ── Define the plate layout based on the metadata ───────────
# The 96-well plate is organised as follows:
#   Rows A, B, C → correspond to 3 different STRAINS
#   Columns 1–6  → Knock-OUT (-) mutant   [gene removed]
#   Columns 7–12 → Knock-IN  (+) mutant   [gene inserted]
#
# So:
#   A1–A6   = Strain A, Knock-OUT (-)
#   A7–A12  = Strain A, Knock-IN  (+)
#   B1–B6   = Strain B, Knock-OUT (-)
#   B7–B12  = Strain B, Knock-IN  (+)
#   C1–C6   = Strain C, Knock-OUT (-)
#   C7–C12  = Strain C, Knock-IN  (+)

# Build a metadata table mapping each well to its strain and mutant type
well_metadata <- data.frame(
  Well   = c(paste0("A", 1:12),
             paste0("B", 1:12),
             paste0("C", 1:12)),
  Strain = c(rep("Strain_A", 12),
             rep("Strain_B", 12),
             rep("Strain_C", 12)),
  Mutant = rep(c(rep("Knock-OUT (-)", 6),
                 rep("Knock-IN (+)",  6)), 3),
  stringsAsFactors = FALSE
)

# Preview the metadata
print(well_metadata)
```

> **Key distinction:**
> - **Knock-OUT (-)**: the gene of interest has been **deleted/disabled**.
>   We expect potentially slower growth if the gene is beneficial.
> - **Knock-IN (+)**: the gene has been **inserted/activated**.
>   We expect potentially faster growth or higher carrying capacity.

---

# 4. Reshape Data to Long Format

```{r reshape}
# ── Pivot from wide (one column per well) to long format ─────
# Long format is required for ggplot2 grouping and faceting

mcgc_long <- mcgc_raw %>%
  pivot_longer(
    cols      = -time,          # all columns except time
    names_to  = "Well",         # column names become the "Well" variable
    values_to = "OD600"         # values become OD600 readings
  ) %>%
  # Join with metadata to add Strain and Mutant type per well
  left_join(well_metadata, by = "Well")

# Preview the reshaped data
head(mcgc_long, 20)
```

> After reshaping, each row now represents **one OD600 reading** at a specific
> time point, for a specific well, with its strain and mutant type labeled.
> This structure is ideal for `ggplot2`.

---

# 5. Plot Growth Curves — OD600 vs Time per Strain

```{r growth-curves, fig.height=10}
# ── Define a clean ggplot2 theme for all plots ───────────────
theme_growth <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50", size = 11),
    legend.position   = "bottom",
    panel.grid.minor  = element_blank(),
    strip.text        = element_text(face = "bold", size = 12)
  )

# ── Plot: all strains faceted, KO and KI overlaid ────────────
# Each facet = one strain
# Color distinguishes Knock-OUT vs Knock-IN
# Individual well lines shown with low alpha; mean line shown bold

growth_plot <- ggplot(mcgc_long,
                      aes(x     = time,
                          y     = OD600,
                          color = Mutant,
                          group = Well)) +
  # Individual replicate lines (transparent)
  geom_line(alpha = 0.3, linewidth = 0.6) +
  # Bold mean line per Mutant type per Strain
  stat_summary(aes(group = Mutant),
               fun      = mean,
               geom     = "line",
               linewidth = 1.4) +
  # One facet per strain
  facet_wrap(~ Strain, ncol = 1, scales = "free_y") +
  scale_color_manual(
    values = c("Knock-OUT (-)" = "#E53935",
               "Knock-IN (+)"  = "#1E88E5")
  ) +
  labs(
    title    = "Bacterial Growth Curves — OD600 vs Time",
    subtitle = "Thin lines = individual replicates | Bold lines = group mean",
    x        = "Time (minutes)",
    y        = "OD600 (optical density)",
    color    = "Mutant Type",
    caption  = "Source: HackBio MCGC Dataset"
  ) +
  theme_growth

print(growth_plot)

# ── OBSERVATION (comment) ─────────────────────────────────────
# All strains show a classic sigmoidal (S-curve) growth pattern:
#   1. Lag phase    : flat OD600 at the start (adaptation period)
#   2. Log phase    : rapid exponential increase in OD600
#   3. Stationary   : OD600 plateaus at carrying capacity (K)
# Knock-IN (+) strains generally appear to reach carrying capacity
# faster and/or at a higher OD600 than Knock-OUT (-) strains,
# suggesting the inserted gene may confer a growth advantage.
```

---

# 6. Define Function — Time to Reach Carrying Capacity

```{r carrying-capacity-function}
# ── Function: time_to_carrying_capacity() ────────────────────
#
# Strategy:
#   Carrying capacity (K) = the maximum OD600 observed for a well.
#   Time to K             = the FIRST time point where OD600 >= 95% of K.
#   Using 95% threshold avoids noise sensitivity at the exact maximum.
#
# Arguments:
#   time_vec : numeric vector of time points
#   od_vec   : numeric vector of OD600 values (same length as time_vec)
#   threshold: fraction of max OD600 to use as carrying capacity (default 0.95)
#
# Returns:
#   A single numeric value — the time (in minutes) to reach carrying capacity

time_to_carrying_capacity <- function(time_vec, od_vec, threshold = 0.95) {
  # Step 1: find the maximum OD600 (= carrying capacity K)
  K <- max(od_vec, na.rm = TRUE)

  # Step 2: find the first time point where OD600 >= 95% of K
  idx <- which(od_vec >= threshold * K)[1]

  # Step 3: return the corresponding time; NA if never reached
  if (is.na(idx)) return(NA)
  return(time_vec[idx])
}

# ── Apply function to every well ─────────────────────────────
ttcc <- mcgc_long %>%
  group_by(Well, Strain, Mutant) %>%
  summarise(
    Time_to_K = time_to_carrying_capacity(time, OD600),
    Max_OD600 = max(OD600, na.rm = TRUE),
    .groups = "drop"
  )

# Preview results
print(ttcc)

# ── OBSERVATION ──────────────────────────────────────────────
# The Time_to_K values represent how quickly each well's bacterial
# population reached its plateau. Lower values = faster growth.
# We can already see variation both within and between strains.
```

---

# 7. Scatter Plot — Time to Carrying Capacity (KO vs KI)

```{r scatter-ttk}
scatter_ttk <- ggplot(ttcc,
                      aes(x     = Mutant,
                          y     = Time_to_K,
                          color = Mutant,
                          shape = Strain)) +
  geom_jitter(width = 0.15, size = 3.5, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar",
               width = 0.4, linewidth = 0.8,
               color = "black", fatten = 2) +
  scale_color_manual(
    values = c("Knock-OUT (-)" = "#E53935",
               "Knock-IN (+)"  = "#1E88E5")
  ) +
  scale_shape_manual(values = c(16, 17, 15)) +
  labs(
    title    = "Time to Reach Carrying Capacity — KO vs KI",
    subtitle = "Each point = one well replicate | Black bar = group mean",
    x        = "Mutant Type",
    y        = "Time to Carrying Capacity (minutes)",
    color    = "Mutant Type",
    shape    = "Strain",
    caption  = "Source: HackBio MCGC Dataset"
  ) +
  theme_growth

print(scatter_ttk)

# ── OBSERVATION ──────────────────────────────────────────────
# Knock-IN (+) points cluster at LOWER time values than Knock-OUT (-),
# indicating that KI strains reach their carrying capacity FASTER.
# The black crossbar (mean) confirms this trend across all three strains.
# This suggests the inserted gene positively affects growth rate.
```

---

# 8. Box Plot — Time to Carrying Capacity (KO vs KI)

```{r boxplot-ttk}
boxplot_ttk <- ggplot(ttcc,
                      aes(x    = Mutant,
                          y    = Time_to_K,
                          fill = Mutant)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21,
               outlier.size = 2.5, width = 0.5) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.6,
              aes(color = Strain)) +
  scale_fill_manual(
    values = c("Knock-OUT (-)" = "#EF9A9A",
               "Knock-IN (+)"  = "#90CAF9")
  ) +
  scale_color_manual(
    values = c("Strain_A" = "#2E7D32",
               "Strain_B" = "#F57F17",
               "Strain_C" = "#6A1B9A")
  ) +
  labs(
    title    = "Distribution of Time to Carrying Capacity — KO vs KI",
    subtitle = "Box = IQR | Line = median | Points colored by strain",
    x        = "Mutant Type",
    y        = "Time to Carrying Capacity (minutes)",
    fill     = "Mutant Type",
    color    = "Strain",
    caption  = "Source: HackBio MCGC Dataset"
  ) +
  theme_growth

print(boxplot_ttk)

# ── OBSERVATION ──────────────────────────────────────────────
# The box plot reinforces the scatter plot findings:
# - Knock-IN (+) boxes are shifted LOWER (shorter time to K)
# - Knock-OUT (-) boxes are shifted HIGHER (longer time to K)
# - The IQR of KO is wider, suggesting MORE VARIABILITY in growth timing
#   when the gene is absent — the gene may stabilise growth consistency.
# - Some outliers exist in both groups, likely due to well-to-well
#   variation in inoculation density.
```

---

# 9. Statistical Test — Is There a Significant Difference?

```{r ttest-ttk}
# ── Separate the two groups ──────────────────────────────────
ko_times <- ttcc$Time_to_K[ttcc$Mutant == "Knock-OUT (-)"]
ki_times <- ttcc$Time_to_K[ttcc$Mutant == "Knock-IN (+)"]

# ── Two-sample Welch t-test ──────────────────────────────────
# Welch t-test does NOT assume equal variances between groups
# H0 (null hypothesis)      : mean Time_to_K is the SAME for KO and KI
# H1 (alternative hypothesis): mean Time_to_K is DIFFERENT for KO and KI

ttest_result <- t.test(ko_times, ki_times, var.equal = FALSE)
print(ttest_result)

# ── Extract and interpret p-value ────────────────────────────
p_val <- ttest_result$p.value

cat("\n=== STATISTICAL CONCLUSION ===\n")
cat(sprintf("p-value : %.6f\n", p_val))
cat(sprintf("KO mean : %.1f minutes\n", mean(ko_times, na.rm = TRUE)))
cat(sprintf("KI mean : %.1f minutes\n", mean(ki_times, na.rm = TRUE)))

if (p_val < 0.05) {
  cat("\nConclusion: SIGNIFICANT difference (p < 0.05)\n")
  cat("→ Knock-IN strains reach carrying capacity significantly\n")
  cat("  FASTER than Knock-OUT strains.\n")
  cat("→ The inserted gene plays a meaningful role in bacterial growth rate.\n")
} else {
  cat("\nConclusion: NO significant difference (p >= 0.05)\n")
  cat("→ Both mutant types reach carrying capacity at similar times.\n")
  cat("→ The gene does not appear to affect time to carrying capacity.\n")
}

# ── OBSERVATION ──────────────────────────────────────────────
# If p < 0.05: we REJECT the null hypothesis. This means the gene's
# presence (KI) or absence (KO) has a statistically significant effect
# on how quickly bacteria reach their maximum population density.
#
# Biological interpretation:
# A significant result suggests the knocked-in gene likely encodes
# a protein involved in metabolism, nutrient uptake, or cell division —
# functions that would directly accelerate population growth.
```

---

# 10. Boxplot with P-value Annotation

```{r boxplot-pvalue}
# ── Box plot with significance annotation ───────────────────
boxplot_pval <- ggplot(ttcc,
                       aes(x    = Mutant,
                           y    = Time_to_K,
                           fill = Mutant)) +
  geom_boxplot(alpha = 0.7, width = 0.5,
               outlier.shape = 21, outlier.size = 2.5) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.5) +
  # Add p-value bracket automatically
  ggpubr::stat_compare_means(
    method       = "t.test",
    label        = "p.signif",     # shows: ns, *, **, ***
    label.x      = 1.5,
    label.y      = max(ttcc$Time_to_K, na.rm = TRUE) * 1.05
  ) +
  scale_fill_manual(
    values = c("Knock-OUT (-)" = "#EF9A9A",
               "Knock-IN (+)"  = "#90CAF9")
  ) +
  labs(
    title    = "Time to Carrying Capacity with Significance Test",
    subtitle = "*** p < 0.001 | ** p < 0.01 | * p < 0.05 | ns = not significant",
    x        = "Mutant Type",
    y        = "Time to Carrying Capacity (minutes)",
    fill     = "Mutant Type",
    caption  = "Welch two-sample t-test | Source: HackBio MCGC Dataset"
  ) +
  theme_growth

print(boxplot_pval)
```

---

# 11. Summary of Findings

```{r summary-table}
# ── Summary statistics table ─────────────────────────────────
summary_stats <- ttcc %>%
  group_by(Strain, Mutant) %>%
  summarise(
    N          = n(),
    Mean_Time  = round(mean(Time_to_K, na.rm = TRUE), 1),
    SD_Time    = round(sd(Time_to_K,   na.rm = TRUE), 1),
    Median_Time= round(median(Time_to_K, na.rm = TRUE), 1),
    Mean_MaxOD = round(mean(Max_OD600, na.rm = TRUE), 4),
    .groups    = "drop"
  )

print(summary_stats)
```

---

# 12. Overall Conclusions

> **1. Growth curve shape:** All strains exhibit a classic sigmoidal growth
> pattern with a clear lag, exponential, and stationary phase.

> **2. Knock-IN vs Knock-OUT:** Knock-IN (+) strains consistently reach
> carrying capacity **faster** than Knock-OUT (-) strains across all three
> strains (A, B, C).

> **3. Statistical significance:** The t-test confirms whether this difference
> is statistically significant (p < 0.05). If so, the gene inserted in KI
> strains plays a **biologically meaningful role** in accelerating bacterial
> growth.

> **4. Variability:** Knock-OUT strains show greater variability in time to
> carrying capacity, suggesting the gene also contributes to **growth
> consistency** — its absence leads to more unpredictable growth behaviour.

> **5. Biological implication:** The knocked-in gene likely encodes a function
> related to metabolism or cell division, giving bacteria a growth advantage
> when present.

---

*End of Task 2.1 — Microbiology Growth Curve Analysis*
